---
title: 'DATA605: Discussion Post Week 7'
author: "Andrew Bowen"
date: "2023-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(dplyr)
```

## DATA605: Discussion Post Week 7
***Exercise 3.2.6**: A die is rolled twice. Let X denote the sum of the two numbers that turn up, and Y the difference of the numbers (specifically, the number on the first roll minus the number on the second). Show that E(XY ) = E(X)E(Y ). Are X and Y independent?*

Let's define some global variables:
```{r global-vars}
omega <- expand.grid(1:6, 1:6)
n_omega <- nrow(omega)
face_vals <- seq(1:6)

# Getting all combinations of rolls, as well as their sums & diffs
combos <- omega %>%
            rename(die1=Var1) %>%
            rename(die2=Var2) %>%
            mutate(sum=die1 + die2) %>%
            mutate(diff=die1 - die2)
```

First, let's calculate the expected value of $X$, our sum of the two numbers of the dice roll. We define our sample space for the random variable $X$ to be:

\begin{aligned}
  \Omega_X = \{x | 2 \le x \le 12\}
\end{aligned}

Let's calculate this expected value (`exp_x`) with the below R code. We'll generate our sampel space $\Omega_X$ via the [built-in R function `expand.grid`](https://www.tutorialspoint.com/how-to-create-sample-space-of-throwing-two-dices-in-r)
```{r exp-X}
# First, let's loop over the possible values of X
omega_x <- range(2:12)

# Getting counts (frequencies of each sum in our sample space)
sum_counts <- combos %>%
              count(sum)

exp_x <- 0
for (x in 2:12){
  # Get count of outcomes with given sum
  count_x <- sum_counts %>%
              filter(sum == x) %>% 
              pull(n)
  
  # Calculate probability of that outcome, add to expected value sum
  prob <- x * (count_x / n_omega)
  exp_x <- exp_x + prob
}

print(exp_x)
```


Next, let's define our sample space for Y. This is the possible values that the random variable Y could take as the *difference* of our two dice rolls

\begin{aligned}
  \Omega_Y = \{x | -5 \le x \le 5\}
\end{aligned}

We can take a similar approach as above, but with a slightly different sample space
```{r exp-Y}
# First, let's loop over the possible values of X
omega_y <- range(-5:5)

# Getting counts (frequencies of each sum in our sample space)
diff_counts <- combos %>%
                count(diff)

exp_y <- 0
for (y in -5:5){
  # Get count of outcomes with given sum
  count_y <- diff_counts %>%
              filter(diff == y) %>% 
              pull(n)
  
  # Calculate probability of that outcome, add to expected value sum
  prob <- y * (count_y / n_omega)
  exp_y <- exp_y + prob
}

print(exp_y)
```

So based on the above, we get $E(X)E(Y) = 7 \cdot 0 = 0$ as the value for $E(Y)$ above is 0 due to a floating point error.

```{r}
(exp_x * exp_y)
```

Now we need to calculate $E(XY)$ and show that it is equal to zero to complete our proof. We can do a similar method above, using the sample space of the product of X and Y as our "new" random variable over which we can calculate the expected value.
```{r exp-value}
# Let's define our product of X and Y column first in our combos df
combos <- combos %>%
  mutate(XY = sum * diff)

omega_xy <- min(combos['XY']) : max(combos['XY'])

exp <- 0
for (i in omega_xy){
  # Get count of outcomes with given sum
  cnt <- diff_counts %>%
              filter(diff == i) %>% 
              pull(n)
  
  # Calculate probability of that outcome, add to expected value sum
  prob <- i * (cnt / n_omega)
  exp <- exp + prob
}

print(exp)
```

We're seeing here that $E(XY) =0$ as well, which holds from our result above for $E(X)E(Y) = 0$. The random variables X and Y are independent, since the roll of each die is independent from each other. In other words, you can't predict Y from the outcome of X alone, and vice versa, so X and Y are indpenent random variables. Also, the fact that their expected values observe this theorem adds to the case of them being independent.