---
title: 'DATA605: Discussion Week 13'
author: "Andrew Bowen"
date: "2023-04-24"
output: html_document
---
```{r}

```

## Exercise 4.2.15 - Related Rates
15. A company that produces landscaping materials is dumping sand into a conical pile. The sand is being poured at a rate of $5 ft^3/s$; the physical properties of the sand, in conjunction with gravity, ensure that the cone’s height is roughly $2/3$ the length of the diameter of the circular base.
How fast is the cone rising when it has a height of 30 feet?

Let's take the cone's height to be $h(t)$, and the diameter of the conde to be $d = 2r(t)$, where $r$ is the radius of the cone. Finally, let the time-dependent volume $V(t)$ of the cone equal:

\begin{aligned}
  V(t) = \frac{\pi}{3}hr^2
\end{aligned}

Where we know

\begin{aligned}
  \frac{dV}{dt} = 5
\end{aligned}

First, let's write a function to calculate the volumne of a cone, based on its radius `r` and height `h`
```{r}
# First, define a function for the volume of the cone, V(t)
V <- function(r, h){
  return((pi /3) * (r^2) * h)
}

# Test call of volumne function
261.7994 == round(V(5, 10), 4)
```

Now, to solve this related rates problem, we'll need to evaluate the rate of change (wrt time $t$) of the sand pile's height, when $h(t) = 30$. In other words, we need to evaluate the following expression:

\begin{aligned}
  \frac{dh}{dt} \rvert_{h=30}
\end{aligned}


As a simplifying assumption, let's assume $h(t)$ is a monotonically increasing function of $t$, meaning sand is **not** taken away from the pile over our timeframe of interest.

Knowing the relation between our conical height and radius: $h = \frac{2}{3}d = \frac{4}{3}r$, we can substitute to get $V$ as an expression of just $h$:

\begin{aligned}
  V(t) = \frac{\pi}{3}hr^2 \newline
       = \frac{1}{3}\pi(\frac{3}{4}h)h \newline
       = \frac{\pi}{4}h^2
\end{aligned}

Alternatively, we can write R functions to compute $V$ and $h$ as functions of each other
```{r}
# rewritten function of volume as a function of height h
V_h <- function(h){
  return(pi * (1/16) * h^3)
}

h_V <- function(V){
  # Height as a function of volume (both wrt to t)
  (V * 16 / pi)^(1/3)
}
```

We don't know $h$ as a function of time explicitly, but can manipulate differentials to get to where we want to go, much like 

\begin{aligned}
  \frac{dh}{dt} = \frac{dV}{dt} \cdot \frac{dh}{dV} \newline
                = 5 \cdot \frac{dh}{dV}
\end{aligned}

Now we can differentiate our expression for $h$ with respect to $V$:

\begin{aligned}
  \frac{dh}{dV} = \frac{d}{dV}\left ( \frac{4V^{\frac{1}{3}}}{\sqrt{\pi}} \right )\newline
                = \frac{4}{3\sqrt{\pi}V^{\frac{2}{3}}}
\end{aligned}

Now we just need to evaluate when $h(t) = 30$, using our relations for $h, r, V$ from above

\begin{aligned}
  \frac{dh}{dt}\rvert_{h=30} = \frac{5 \cdot 2}{\sqrt{\pi V\rvert_{h=30}}}
\end{aligned}

Let's calculate the volume when the height of our conical pile is $30ft$ and plug in:

\begin{aligned}
  V\rvert_{h=30}(t) = \frac{\pi h}{3}(\frac{90}{4})^{2}\newline
                    \approx 15904.31
\end{aligned}

Let's verify our volume calculation in R:
```{r}
r <- 90/4
h <- 30
(v30 = V(r, h))
```

Plugging into our above expression yields:

\begin{aligned}
  \frac{dh}{dt}\rvert_{h=30} = \frac{5 \cdot 2}{\sqrt{\pi V\rvert_{h=30}}} \newline
                             = \frac{10}{3 \sqrt{\pi}(12904.31)^{\frac{2}{3}}}\newline
                             \approx 0.003
\end{aligned}

We can verify the above calculation in R
```{r}
(dhdV <- (10 / (3 * sqrt(pi) * (v30^(2/3)))))
```
In plain english, the height of our conical pile is changing at about $0.003 ft/s$ when the height of the cone is 30ft.

## Solving using R
We can use the base R functions `deriv` to calculate the derivative, then evaluate this expression at $h(t) = 30$ to ensure we did the math right
```{r}
# function for height with volume as a param
v_exp <- (10 / (3 * sqrt(pi) * (v30^(2/3))))

# Call derivative to find dh/dt at h(t) = 30ft
dh_dV <- deriv(h ~v_exp, c("V","h"), func=TRUE)

# Call our derivative function with our volume value at h=30 as a param
dh_dV(v30)[1] == dhdV
```

