---
title: 'DATA605: Discussion Post Week 11'
author: "Andrew Bowen"
date: "2023-04-04"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Dataset
For this exercise, I chose the `faithful` dataset which is available as an R dataset


First, let's load the built-in `cars` dataset to a native R dataframe
```{r}
df <- as.data.frame(faithful)

head(df)
```

# Model Creation
Let's use the built-in `lm` function from R
```{r linear-model}
# creating our model of eruption duration as a function of wait time
model <- lm(faithful$eruptions ~ faithful$waiting)

summary(model)
```

# Model Analysis
Let's plot our data as well as our model line
```{r}
plot(faithful$waiting, faithful$eruptions)
abline(model)
```
There appears to be a general linear relationship between these variables in this plot. We should continue to analyze our assumptions for regression.

Next, let's plot the distribution of our residuals via a histogram
```{r}
res <- resid(model) #alternatively, model$residuals
hist(res)
```
This distribution is slightly skewed, but looks to be relatively normal from an eye test

Next let's plot our residuals as a function of our predicted values
```{r}
plot(fitted(model), res)
```
There appears to be a slight pattern in these residuals, as there looks to be two clusters of data within our residuals plot

Lastly, we can create a Q-Q plot to evaluate the normality of our residuals.
```{r}
# Create Q-Q plot of our residuals
qqnorm(res)
qqline(res)
```


While this is relatively linear as a Q-Q plot, the tails of the distribution wander away from our normalization line. Overall, a linear model fits the data and meets the assumptions of regression, but the clustering behavior displayed above leads me to believe a different model may be appropriate for this dataset.


