---
title: "DATA605 Discussion Post Week 10"
author: "Andrew Bowen"
date: "2023-03-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 11.1.13
*Write a program to compute $u^{(n)}$ given $\textbf{u}$ and $P$. Use this program to compute $u^{(10)}$ for the Land of Oz example, with $u = (0,1,0)$, and with $u = (1/3, 1/3, 1/3)$.*


First, let's define our transition probability matrix $P$ for each weather type (*Rainy, Sunny, Nice*)
```{r}
# First, define our P matrix
n <- 10
oz_matrix <- matrix(c(0.5, 0.25, 0.25, 0.5, 0, 0.5, 0.25, 0.25, 0.5), nrow=3, ncol=3, byrow=TRUE)

rownames(oz_matrix) <- c("R", "N", "S")
colnames(oz_matrix) <- c("R", "N", "S")

print(oz_matrix)
```


Based on Theorem 11.2, if $P$ is our transition matrix, Then the probability that the chain is in state $s_i$ after $n$ steps is the $i^{th}$ entry in the vector:
$$
\begin{aligned}
  \textbf{$u^{(n)}$} = \textbf{u}P^n
\end{aligned}
$$

For the case where $\textbf{u} = (0, 1, 0)$:
```{r}
u <- c(0, 1, 0)
P <- oz_matrix

for (i in 1:n -1){
  P <- P %*% P
}

(x <- u %*% P)
```

For the case where $\textbf{u} = (1/3, 1/3, 1/3)$:
```{r}
u <- c(1/3, 1/3, 1/3)
P <- oz_matrix

for (i in 1:n - 1){
  P <- P %*% P
}

(y <- u %*% P)
```


Since the result of these calculations are probability vectors, we should be able to sum the distribution vectors to 1. 
```{r}
print(sum(x))
print(sum(y))
```

Ideally, we could wrap the above block of code in a function like:
```{r}
calculate_prob <- function(U, M, N){
  if (n == 1){
    return(U %*% M)
  }
  else{
    for (i in 1:(n)){
      M <- M %*% M
    }

    y <- u %*% M
    return(y)
  }
}

# Test call our function for the above case
calculate_prob(c(0, 1, 0), oz_matrix, 10)
calculate_prob(c(1/3, 1/3, 1/3), oz_matrix, 10)
```
