---
title: 'DATA605: Problem Set 13'
author: "Andrew Bowen"
date: "2023-04-26"
output: html_document
---

# Question 1
Let's pick our substitution function $u = -7x$, so $\frac{\,du}{\,dx} = -7$, meaning $\frac{\,du}{-7} = \,dx$. Plugging in we get
\begin{aligned}
  \int 4e^{-7x}\,dx = 4\int e^{-7x} \,dx\newline
                    = 4\int \frac{1}{-7}e^{u} \,du\newline
                    = \frac{-4}{7}\int e^u \,du \newline
                    = \frac{-4}{7}e^u + C = \frac{-4}{7}e^{-7x} + C
\end{aligned}

# Question 2
We can compute the indefinite integral of $\frac{dN}{dt}$ to get an expression for $N(t)$. From there we can use our boundary condition ($N(1) = 6530$) to find an expression

\begin{aligned}
  N(t) = \int \frac{-3150}{t^4} - 220 \,dx = \frac{1050}{3t^3} - 220x +  C
\end{aligned}

Now we can plug in our boundary condition to find our constant of integration $C$:

\begin{aligned}
  N(1) = 6530 = \frac{1050}{3(1)^3} - 220(1) + C\newline
              = 350 - 220 + C\newline
            C = 6530 - 350 + 220 
\end{aligned}
```{r biology}
C <- 6530 - 350- 220

print(C)
```
Making our final expression for the number of bacteria per day: $N(t) = \frac{1050}{3t^3} + 5960$
```{r bacteria-check}
N <- function(t, c=C){
  1050/(3* t^3) + c
}

(N_prime <- deriv(expression(N), "t"))
```

# Question 3
```{r}
f <- function(x){
  2 * x - 9
}

x_vals <- c(5, 6, 7, 8)

# Iterate over rectangles (width = 1). and sum up area of riemann rectangles
area <- 0 
for (x in x_vals){
  rect_area <- 1 * f(x)
  area <- area + rect_area
}

print(area)
```


# Question 4
Let's define our functions in question as:

\begin{aligned}
  g(x) = x + 2\newline
  h(x) = x^2 - 2x - 2
\end{aligned}

We can solve this by finding where these functions equal each other ($g(x) = h(x)$), then subtract the definite integral of each function over that interval (assuming only 2 intercepts)
```{r}
g <- function(x){
  x + 2
}
h <- function(x){
  x^2 - 2*x - 2
}

# Plotting our functions
curve(h, from =-5, to=5, col=2)
curve(g, from =-5, to=5,  col=3, add=TRUE)
```

Visually, it looks like these functions intersect at $x = -1, 4$. We can solve this intersect analytically:

\begin{aligned}
  x + 2 = x^2 - 2x - 2\newline
  0 = x^2 - 3x - 4\newline
  0 = (x - 4)(x - 1)
\end{aligned}

Confirming our intuition above, this equation has roots at $x = -1, 4$.

```{r, area-between-curves}
x1 <- -1
x2 <- 4

int_g <- integrate(g, x1, x2)
int_h <- integrate(h, x1, x2)

(area <- int_g[1]$value - int_h[1]$value)
```

# Question 5
Let $x$ be the number of flat irons per order, and $n$ be the number of orders. This means that $nx = 110$. Let's also introduce the below functions for the cost of storage ($S(x)$) and ordering $\Gamma(X)$. 

\begin{aligned}
  S(x) = 3.75x\newline
  \Gamma(x) = 8.25\frac{110}{x}
\end{aligned}

Let the total cost function be $C(x) = S(x) + \Gamma(x) = 3.75x + \frac{907.5}{x}$. To find a minimal cost as a function of flat irons ordered, we'll need to find where $C'(x) = 0$

\begin{aligned}
  C'(x) = 3.75 - \frac{907.5}{x^2}\newline
      0 = 3.75 - \frac{907.5}{x^2}\newline
      3.75 = \frac{907.5}{x^2}
      x = 16.02
\end{aligned}

Now, plugging in for $n$, rounding x to 16

\begin{aligned}
  nx = 110\newline
  n(16) = 110
  n = 6.875
\end{aligned}

# Question 6
First, let's choose our $u$ and $v$ parts of our integral. Let 

\begin{aligned}
  u = \ln(9x)\newline
  v = \frac{x^7}{7}\newline
  \,du = \frac{9}{x}\,dx
\end{aligned}

We can plug into the integration by parts formula:

\begin{aligned}
  \int u\,dv = uv - \int v\,du\newline
             = \ln(9x)\frac{x^7}{7} - \int \frac{x^6}{7}\,dx\newline
             =  \ln(9x)\frac{x^7}{7} -  \frac{x^7}{49} + C
\end{aligned}

# Question 7
We can check if a function over a given integral is a probability density by integrating the function over that integral. If the expression evaluates to 1, we have a probability density function:

```{r pdf}
f <- function(x){
  return(1 / (6 * x))
}

# Now let's integrate over our interval
(i <- integrate(f, 1, exp(6)))
```

Since our integral evaluates to 1, our function $f(x) = \frac{1}{6x}$ is a PDF on the interval $[1, e^6]$

