---
title: 'DATA605: Problem Set 7'
author: "Andrew Bowen"
date: "2023-03-07"
output: html_document
---

```{r, message=FALSE}
library(glue)
```

# Question 1

Since each random variable $X_i$ can take values from $1, 2, \dots k$, and there are $n$ random variables from which Y could be derived, there are $k^n$ possible values for $Y = min(X_1, X2, \dots, X_n)$. Additionally, since we know that each random variable $X_i$ are independent, we can multiply their probabilities.

Since our uniform distribution is over the integers, we can find the probability distribution of $Y = min(X_i)$ in the window between $Y = y$ and $Y = y-1$: $P(Y = y) = P(Y \le y) - P(Y \le y - 1)$


\begin{aligned}
  P(Y \le y) = 1 - P(Y \ge y)
             = 1 - P(min(X_i) \ge y)
             = 1 - \prod_{i = 1}^{n} P(X_i = y)
             = 1 - \frac{(k - y)^n}{k^n}
             
\end{aligned}

We can plug in $y-1$ for $y$ in the above definition to find our distribution for $Y$:

\begin{aligned}
  P(Y \le y - 1) = 1 - \frac{(k - (y - 1))^n}{k^n} = 1 - \frac{(k - y + 1)^n}{k^n} 
\end{aligned}

So we're left with our final distribution for $Y$:

\begin{aligned}
  P(Y = y) = P(Y \le y) - P(Y \le y - 1) = 1 - \frac{(k - y )^n}{k^n} - 1 + \frac{(k - y + 1)^n}{k^n} 
           = \frac{(k - y + 1)^n - (k -y)^n}{k^n} 
\end{aligned}

# Question 2


## 2a

We'll use the geometric distribution `dgeom` here to find the probability of failure after 8 years, given the probability of failing in a given year is 0.1
```{r 2a}
p_fail <- 0.1 # probability of failure in given year
n <- 8

(prob_fail_8 <- dgeom(n, p_fail))
```

Now let's find the [expected value and variance of this geometric distribution](https://en.wikipedia.org/wiki/Geometric_distribution#Properties)
```{r exp-value-var-2a}
exp_value <- 1 / p_fail

v <- (1 - p_fail) / p_fail**2
sigma <- sqrt(v)

print(glue("Expected Value: {exp_value}"))
print(glue("Std Dev: {sigma}"))
```


## 2b - Exponential Distribution
Using the exponential distribution (R's built-in `dexp` function as a probability density), we can model the probability of machine failure after 8 years
```{r exponential}
# Probability of failure after 8 years (exponential)
(p_fail_exp <- dexp(n, rate=p_fail))
```

The [expected value and variance of an exponential distribution](https://en.wikipedia.org/wiki/Exponential_distribution#Properties) is based on the *rate parameter $\lambda$. In this case our average rate of failure for the machine is 1 failure every 10 years, or a rate of 0.1 failures per year

```{r}
exp_value_exp <- 1 / p_fail
var_exp <- 1 / (p_fail ** 2)
std_exp <- sqrt(var_exp)


print(glue("Expected Value (exponential): {exp_value_exp}"))
print(glue("Std Dev (exponential): {std_exp}"))
```


## 2c - Binomial Distribution
Similar to above, we can use R's `dbinom` 
```{r binomial}
(p_fail_binom <- dbinom(0, n, p_fail))
```

The expected value and variance of the binomal distribution can be [found here](https://en.wikipedia.org/wiki/Binomial_distribution#Properties). Assuming $n$ is the number of trials (years) and p is the probability of each trial resulting in a "success" (in this case the machine failing is a "success" with our setup, even though the language may be confusing)


\begin{aligned}
  E(X) = np
\end{aligned}

\begin{aligned}
  Var(X) = npq = np(1 - p)
\end{aligned}


```{r}
# Calculateing expected value, variance and std dev of binomial distribution
exp_value_binom <- n * p_fail
var_binom <- n * p_fail * (1 - p_fail)
std_binom <- sqrt(var_binom)


print(glue("Expected Value (exponential): {exp_value_binom}"))
print(glue("Std Dev: {std_binom}"))
```


## 2d - Poisson Distribution
Again, we can use the built-in `dpois` function to find the probability of failure after 8 years
```{r poisson}
lambda <- 0.1
(p_fail_poisson <- dpois(8, lambda))
```

For poisson distribution, the [expected value and variance are both equal to $\lambda$.](https://en.wikipedia.org/wiki/Poisson_distribution#Properties). In this case that means they are both equal to $\lambda=0.1$ as that is the average rate of machine failure in a given year





