---
title: "DATA605 Homework 2"
author: "Andrew Bowen"
date: "2023-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1

1. *Prove that $A^TA \ne AA^T$ in general*.   
Assume $A$ is an $m x n$ matrix, with $m \in \mathbb{R}$ and $n \in \mathbb{R}$. Let's also assum $m \ne n$. Since A is $m x n$, $A^T$ is then $n x m$. If we just compare dimensions when we multiply each term of the equation, we see that each side could result in differing dimensions. 

$$
\begin{align}
_{(m x n)}AA^T_{(nxm)} = B_{(m x m)}
\end{align}
$$


$$
\begin{align}
_{(n x m)}A^TA_{(mxn)} = C_{(n x n)}
\end{align}
$$
The subscripts in parentheses above denote the dimensions of the corresponding matrices.Because $C$ is $n x n$, and $B$ is $m x m$, and $m$ is not necessarily equal to $n$, then $B \ne C$ and $AA^T \ne A^TA$.


2. *For a special type of square matrix A, we get $A^T A = AA^T$ . Under what conditions
could this be true? (Hint: The Identity matrix I is an example of such a matrix).*

To solve this, let's take an (educated) guess that $A$ is diagonal, which would include the identity matrix $\mathbb{I}$. Then, we can check our assumption and tha $A$ meets the condition

A diagonal matrix $A$ takes the form:


\begin{align}
A = \begin{bmatrix} 
    a_{11} & 0 & \dots \\
    \vdots & \ddots & \\
    0 &             & a_{KK} 
    \end{bmatrix}

\end{align}


with $a_{ii} \ne 0$ and $a_{ij} = 0$ when $i \ne j$. In simpler terms, only the diagonal terms of the matrix are non-zero. When this is the case, $A = A^T$ and we can rewrite the original equation substituting $A$ for $A^T$: 


\begin{align}
  AA^T = AA = A^2
\end{align}

\begin{align}
  A^TA = AA = A^2
\end{align}

Since both sides of the equation are equal to $A^2$, they are equal to each other, and $AA^T = A^TA$ in the case that $A$ is a diagonal matrix.

## `A = LU` Factorization
First, let's define our function(s) to factorize a given square matrix A into lower and upper triangular matrices: [$A= LU$](https://en.wikipedia.org/wiki/LU_decomposition):

```{r}
# create test 3x3 matrix
sample_matrix <- matrix(data=c(1,2,3,5,5,6,7,8,9), nrow=3, ncol=3)
```

Adding a helper functions that will be called in `lu_factorization` to construct the initial $L$ for any beginning matrix size $n$. $L$ starts as the identity matrix then becomes populated with "row multipliers" below the diagonal.
```{r}
baseL <- function(N){
  # Returns "base" lower-triangular matrix: L
  L <- matrix(data=NA, nrow=N, ncol=N)
  # setting diagonal vals to 1 and other vals to 0
  for (i in seq(N)){
    for (j in seq(N))
      # Set diagonal vals to 1
      if (i == j){
        L[i, j] = 1
      }
      # Set other vals to 0
      else{
        L[i, j] = 0
      }
  }
  return(L)
}
```


Now, let's write out our function to perform LU decomposition of a given matrix
```{r factorization-function}

lu_factorization <- function(A){
  # A is square, so n x n by definition
  n <- nrow(A)
  L <- baseL(n)

  # We will derive U from A, so let's set them as equal to start
  U <- A
  # First, iterate over rows
  for (i in 1:n){
    if (i > 1) { # skip first row in matrix - we won't need to row-reduce it!
      
      # nowmiterate over column vals
      for (j in 1:n){
        multiplier <- U[i, j] / U[j, j]
        if (j < i){ # only need to operate/row-reduce on index pairs below the diagonal
          
          # Set L in proper position to have multiplier (below diagonal)
          L[i, j] = multiplier
          
          # Perform row reduction on U
          U[i, ] = U[i, ] - L[i,j] * U[j, ]
        }
      }
    }
  }
  return(list(L, U))
}

# Sample test call of function
result <- lu_factorization(sample_matrix)
L <- result[[1]]
U <- result[[2]]

# print out our results
print(L)
print(U)

print(L %*% U)
```

```{r test-4-x-4}
# Testing out above LU function on sample 4x4 matrix
matrix4x4 <- matrix(c(2,3,1,5,7,2,3,4,1,0,5,4,3,3,4,5), nrow=4, ncol=4)

result4 <- lu_factorization(matrix4x4)

L4 <- result4[[1]]
U4 <- result4[[2]]

print(L4)
print(U4)

print(L4 %*% U4)
```


