---
title: "DATA605 Problem Set 14: Taylor Series"
author: "Andrew Bowen"
date: "2023-05-07"
output: html_document
---

```{r setup, include=FALSE}
library(calculus)
```

### Question 1
Compute the taylor series (to order 4) of:
\begin{aligned}
  f(x) = \frac{1}{1-x}
\end{aligned}

We can use the `taylor` function included in the `calculus` library within R for this. The above functon is valid everywhere except $x=1$
```{r}
f1 <- function(x){ 1 / (1 - x)}

taylor(f1, var="x", order=4)
```



### Question 2
\begin{aligned}
  f(x) = e^x
\end{aligned}

This function is valid everywhere within $(-\infty, \infty)$. We can calulate the Taylor series to n=4
```{r}
# f1 <- function(x){ 1 / (1 - x)}

taylor("exp(x)", var="x", order=4)
```

### Question 3
\begin{aligned}
  f(x) = \log(1 + x)
\end{aligned}
This function is only defined when $x > -1$. We can compute the Taylor series, centered at 0
```{r}
f3 <- function(x){log(1 + x)}

(t3 <- taylor(f3, var=0, order=4))
```


### Question 4
\begin{aligned}
  f(x) = x^{1/2} = \sqrt{x}
\end{aligned}

The above function is only valid and differentiable over the reals when $x > 0$. We can calculate the Taylor series centered at $x=1$, as that's a valid point of reference
```{r}
f4 <- function(x){sqrt(x)}

(t4 <- taylor(f4, var=1, order=4))
```


