---
title: "DATA 605 HW1"
author: "Andrew Bowen"
date: "2023-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sample-code}
x=c(rep(0,500),seq(0,1,length.out=1000), rep(1,500))
y=c(seq(-1,1,length.out=500),rep(0,1000), seq(-1,1,length.out=500))
z=rbind(x,y)
plot(y~x, xlim=c(-3,3), ylim=c(-3,3))
```

First, let's construct an image of my initials (AB)
```{r}
# Each rep in x needs to have a corresponding seq in y (and vice versa)
x=c(# B lines
    rep(0,500),
    seq(0,1,length.out=500),
    rep(1,500),
    seq(0,1,length.out=500),
    seq(0,1,length.out=500),
    
    
    # A lines
    seq(-1.5, -0.5, length.out = 500),
    rep(-1.5, 500),
    rep(-0.5, 500),
    seq(-1.5, -0.5, length.out = 500)
    
    )
y=c(# B lines
    seq(-1,1,length.out=500),
    rep(0,500),
    seq(-1,1,length.out=500),
    rep(1,500),
    rep(-1,500),
    
    
    # A lines
    rep(0,500),
    seq(-1,1, length.out = 500),
    seq(-1,1, length.out = 500),
    rep(1, 500)
    
    
    )
z=rbind(x,y)
plot(y~x, xlim=c(-3,3), ylim=c(-3,3))

```
Looks like we wera able to plot initials. Now let's get to some matrix transforms!

First, let's use the built-in `diag` function to construct an identity matrix of whichever dimension we want ($n x n$)
```{r}
matrix <- diag(2)

matrix
```

## Shear

In a [shear matrix](https://en.wikipedia.org/wiki/Shear_matrix), the off-diagonal values determine the amount of shear


```{r, y-shear}
# Create plotting window. XQuartz needs to be installed locally for this to work
x11()

# Plotting y-shear in loop
for (i in seq(-3,3, length.out=100)){
  # Update the identity matrix to have the shear coefficient in the off-diagonal position
  # For shear in the y-plane, this comes under the diagonal (row 2 col 1)
  matrix[2,1] = i
  # Apply matrix multiplication to z vector and plot results
  sheared_matrix <- apply(z, 2, function(x) matrix %*% x)
  plot(sheared_matrix[2,] ~ sheared_matrix[1,])
}
```

```{r x-shear}
x11()

matrix <- diag(2)
# Plotting x-shear in loop
# Only thing that differs is indices when adjusting shear matrix
for (i in seq(-3,3, length.out=100)){
  # Update the identity matrix to have the shear coefficient in the off-diagonal position
  # For shear in the y-plane, this comes under the diagonal (row 2 col 1)
  matrix[1,2] = i
  
  # Apply matrix multiplication to entire z vector and plot results
  sheared_matrix <- apply(z, 2, function(x) matrix %*% x)
  plot(sheared_matrix[2,] ~ sheared_matrix[1,])
}
```


## Scaling
For scaling, we don't add any off-diagonal entries, but [instead scale the diagonal values to our "scaling factor" (i.e., for a scale of 2 we would multiply the diagonal values by 2 each)
```{r scaling}
x11()
identity_matrix <- diag(2)

for (i in seq(0, 3, length.out=100)){
  # Scaling up our transformation matrix
  scale_matrix <- identity_matrix * i
  
  # Multiplying letter vectors by scale matrix
  scaled_vector <- apply(z, 2, function(x) scale_matrix %*% x)
  plot(scaled_vector[2,] ~ scaled_vector[1,], xlim=c(-3,3), ylim = c(-3,3))
}
```

## Rotation
[Rotation matrices have $sine$ and $cosine$ entries of some specified angle $\theta$](https://en.wikipedia.org/wiki/Rotation_matrix)
```{r rotation}
x11()
for (theta in seq(0, 3, length.out=100)){
  # Creating rotation matrix with parameterized angle theta
  rotation_matrix <-  matrix(c(cos(theta), -sin(theta),
                               sin(theta), cos(theta)), nrow = 2, ncol = 2)
  
  # Multiplying letter vectors by rotation matrix and plotting
  rotated_vector <- apply(z, 2, function(x) rotation_matrix %*% x)
  plot(rotated_vector[2,] ~ rotated_vector[1,], xlim=c(-3,3), ylim = c(-3,3))
}
```

## Projection
[Projection matrices](https://en.wikipedia.org/wiki/Projection_(linear_algebra)) essentially flatten a vector in one-direction, on

```{r projection}
x11()
projection_matrix <- diag(2)
for (i in seq(3, 0, length.out=100)){
  # Updating 
  projection_matrix[2,2] = i
  
  # Multiplying letter vectors by projection matrix and plotting
  projection <- apply(z, 2, function(x) projection_matrix %*% x)
  plot(projection[2,] ~ projection[1,], xlim=c(-3,3), ylim = c(-3,3))
}
```
