---
title: 'DATA605 Final: Kaggle Submission'
author: "Andrew Bowen"
date: "2023-04-30"
output: html_document
---

```{r libraries, echo=FALSE, message=FALSE}
library(tidyverse)
library(MASS)
library(car)
library(lmtest)
library(matrixcalc)
```

### Load our Datasets
I uploaded the kaggle datasets to my GitHub, reading them here for reproducability.
```{r}
test_url <- "https://raw.githubusercontent.com/andrewbowen19/computationalMath605/main/data/test.csv"
train_url <- "https://raw.githubusercontent.com/andrewbowen19/computationalMath605/main/data/train.csv"

test <- read.csv(test_url)
train <- read.csv(train_url)
```

## Descriptive and Inferential Statistics

```{r}
summary(train)
```

Creating a scatter plot of the basement square footage and sale price ($)
```{r}
ggplot(train, aes(x=TotalBsmtSF, y=SalePrice)) + geom_point() + xlab("Basement Square Footage") + ylab("Sale Price ($)")
```

Creating another scatter plot between the lot area ($ft^2$) and `SalePrice`

```{r}
ggplot(train, aes(x=LotArea, y=SalePrice)) + geom_point() + xlab("Lot Area (ft^2)") + ylab("Sale Price ($)")
```

Generating a correlation matrix between our three quantitative variables above (`LotArea`, `TotalBsmtSF`, `SalePrice`)
```{r}

(corTrain <- cor(train[,c("LotArea", "TotalBsmtSF", "SalePrice") ]))
```
Running `cor.test` for each permutation of quantitative variables above
```{r}
for (i in colnames(corTrain)){
  for (j in colnames(corTrain)){
    if (i != j){
      ctest <- cor.test(corTrain[,c(i)], corTrain[,c(j)], conf.level=0.8)
      print(ctest)
    }
  }
}
```


For each combination of variables tested, we have p-values greater than our significance level ($\alpha = 0.2$). This indicates that we can not reject our null hypothesis that the true population correlation between variables is equal to 0. 

### Linear Algebra and Correlation
```{r}
det(corTrain)
```

Our determinant is **not** equal to 0, so we should be able to invert our correlation matrix. We can do this using the built-in `solve` function in R:
```{r invert-cor-matrix}
(invCor <- solve(corTrain))
```
```{r inv-matrix-multiply}
corTrain %*% invCor
```
Accounting for some floating point errors, the above matrix product is the identity matrix, which holds for the multiplication of a matrix $A$ with its inverse $A^{-1}$: $I = AA^{-1}$


LU Decomposition of our correlation matrix from above using the `matrixcalc` package
```{r LU deomp}
(LU <- lu.decomposition(corTrain))
```
Let's verify that our LU decomposition does in fact equal our original correlation matrix when multiplied together: $A = LU$
```{r lu-verification}
A <- LU$L %*% LU$U 
A == corTrain
```

### Calculus-Bases Probability & Statistics
Let's use the `MasVnrArea` variable, which is definitely skew-right. We'll need to remove a few `NA` values from our data first
```{r}
df <- train %>% filter(!is.na(MasVnrArea))
qplot(df$MasVnrArea)
```

Calling `fitdistr` from the `MASS` package to fit our dataset variable to an exponential
```{r}
# Fit our distribution to an exponential
fit <- fitdistr(df$MasVnrArea, "exponential")

# Getting the rate poarameter lambda
lambda <- fit$estimate[["rate"]]
```

Now let's generate random values with our `lambda` variable as the rate parameter
```{r}
dat <- rexp(1000, lambda)
randValues <- as.data.frame(dat, col.names=character(1))


hist(randValues[,1], col=rgb(0,0,1,0.5))
hist(df$MasVnrArea, add=TRUE, col=rgb(1,0,0,0.5))

```

These both look to be skew-right with similar shapes (although the Kaggle data has a bit longer tail). Overall, these historgrams are similarly shaped and likely would have been pulled from the same distribution


**Using the exponential pdf, find the 5th and 95th percentiles using the cumulative distribution function (CDF).**
```{r}
p5 <- pexp(0.05, rate=lambda)
p95 <- pexp(0.95, rate=lambda)
```

Let's now [construct a 95% confidence interval](https://www.r-bloggers.com/2021/11/calculate-confidence-intervals-in-r/) from our Kaggle variable `MasVnrArea`. To do this we'll need our sample mean and standard deviation

```{r conf-interval}
n <- length(df$MasVnrArea)
mu <- mean(df$MasVnrArea)
sigma <- sd(df$MasVnrArea)

margin <- qt(0.975,df=n-1)*sigma/sqrt(n)

lower_interval <- mu - margin
upper_interval <- mu + margin
print(lower_interval)
print(upper_interval)
```

Getting 5th and 95th percentile
```{r}
quantile(df$MasVnrArea, c(0.05, 0.95))
```



### Modeling

```{r}
mylm <- lm(SalePrice ~ FullBath + LotArea + OverallCond + GarageArea, train)

summary(mylm)
```

```{r}
plot(mylm)
```

#### Linear Regression Assumptions

1. Linear Relationship
It doesn't appear that our residuals have any relationship to our fitted values, per our first model plot above.

2. Residuals are distributed normally
We can check this assumption with our QQ-plot. The plot is generally linear in the middle of our residuals, but we can test this with a simple histogram and a [Shapiro Test](https://en.wikipedia.org/wiki/Shapiro–Wilk_test)
```{r}
hist(mylm$residuals)
```

```{r}
shapiro.test(mylm$residuals)
```
Our p-value being less than a significance level of $\alpha=0.05$ indicates we can assume normality of our residuals.


3. Homoscedasticity
We can test this with a [Breush-Pagan test](http://math.furman.edu/~dcs/courses/math47/R/library/lmtest/html/bptest.html) in R
```{r}
bptest(mylm)
```
Our p-value being less than $\alpha=0.05$ means that we can reject the null hypothesis and claim that our variables meet the criterion of homoscedasticity.


### Results
```{r}
test$SalePrice = predict(mylm, newdata = test)
```

Now we can write our predicted values to a csv file 
```{r}
# Get DF of only Id and price column for submission
df <- as.data.frame(test[, c("Id", "SalePrice")])

# impute any missing values for Kaggle as the mean sale price
x <- mean(df$SalePrice, na.rm=TRUE)
df[is.na(df$SalePrice), "SalePrice"] <- x

# Write data to csv
write.csv(df, "submission.csv", row.names=F)
```



Test data & predicted values for sale price regression on home purchase datasets. Final Project for CUNY DATA605
